---
title: "**Lab Program 1**"
author: "**R V Abhishek**"
date: "2025-08-13"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Program to check what type of Triangle given 3 sides, and calculate its area

## Validating the triangle

```{r}
is_valid_triangle <- function(a, b, c) {
  return ((a + b > c) & (b + c > a) & (a + c > b ))
}
```

## Function to check the type of triangle

```{r}
triangle_type <- function(a , b , c) {
  if (a == b && b == c) {
    return(" Equilateral ")
  } else if ( a == b || b == c || a == c) {
    return(" Isosceles ")
  } else {
    return("Scalene")
  }
}
```

## Calculating Area using Heron's Formula

```{r}
triangle_area <- function(a , b , c) {
  s <- (a + b + c) / 2 # Semi - perimeter
  # Heron ’s formula
  area <- sqrt (s * (s - a) * (s - b) * (s - c))
  return (area)
}
```

## Validating inputs

```{r}
validate_input <- function(x) {
  if (!is.numeric(x) || x <= 0) {
    stop("Error : Input must be a positive number.")
  }
  return(TRUE)
}
```

## Main Code Block

1.  Defining 3 variables representing the 3 sides of the triangle

```{r}
cat("Enter the lengths of the sides of the triangle :\n") 
a <- as.numeric(readline(prompt = "Side A: ")) 
b <- as.numeric(readline(prompt = "Side B: ")) 
c <- as.numeric(readline(prompt = "Side C: "))
```

2.  Input Validation and implementation of all the functions.

```{r}
# Input validation} 
tryCatch ({ 
  validate_input(a) 
  validate_input(b) 
  validate_input(c) 
  
  # Check if the inputs form a valid triangle 
  if (!is_valid_triangle(a , b , c)) { 
    stop("Error : The given sides do not form a valid triangle.")
  }
  
  # Determine the type of triangle 
  type_of_triangle <- triangle_type(a , b , c) 
  cat("The triangle is:", type_of_triangle, "\n") 
  # Calculate the area of the triangle 
  area_of_triangle <- triangle_area(a, b, c) 
  cat("The area of the triangle is:", area_of_triangle, "\n")

}, error = function(e){ 
  cat(e$message, "\n") 
})
```

# Sample Output

**Enter the lengths of the sides of the triangle:**

**Side a:** 5

**Side b:** 5

**Side c:** 8

**The triangle is:** Isosceles

**The area of the triangle is:** 12

**Enter the lengths of the sides of the triangle:**

**Side a:** 1

**Side b:** 2

**Side c:** 8

**Error:** The given sides do not form a valid triangle.
